Assessing primitives performance\\ on multi-stage execution

### 2018-05-08 paper.tex

@@

\paragraph{Final specification.} The final specification we productized last month takes two or three parameters. The third parameter, the order, is optional. If the third parameter is absent, the VM assumes the byte ordering (ASCII order) is the order to compare against. Therefore, the VM does not need to go through the indirection order array in the common case. The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.
\cba{Maybe rewrite the last paragraph to make it clearer / more formal. - CF Java spec ? => This is critical, maybe in bold or in a box}

@@

\paragraph{Baseline: pure Smalltalk.}
: baseline
Smalltalk code on top of VM. Used as our baseline for benchmarks.

\begin{code}
compareWithPureST: aString

	| c1 c2 length1 length2 |
	length1 := self size.
	length2 := aString size.
		1 to: (length1 min: length2) do: [ :i |
			(c1 := self basicAt: i) = (c2 := aString basicAt: i) 
			ifFalse: [
				c1 < c2
					ifTrue: [ ^1 ]
					ifFalse: [ ^3 ] ] ].
		length1 = length2 ifTrue: [ ^2 ].
		length1 < length2
			ifTrue: [ ^1 ]
			ifFalse: [ ^3 ] 
\end{code}

Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:
New spec

\paragraph{Existing: Smart Syntax Plugin.}
Restricted Smalltalk

written in Smalltalk with restrictions.
Compiled to C AOT into a plugin.
Write once, fall-back + compiled.

old spec

\paragraph{Existing: Slang Plugin.}
Slang, but no access au fonctions de la VM drectement.
 interpreterProxy. (C compiler inlining, linking time ? cmacro ?)

 old spec

@@

One or the other implementation could be chosen over another one, depending on what one's aiming at: the main trade-off will be between the engineering time and the performance. The table \ref{cmp_implem} summarizes the main pros and cons related to each implementation

@@

This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors. On the contrary, LLVM generates more instructions: it especially fails at perserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.

@@

%% Bibliography
\bibliography{sista}

>>>>

\cba{We need to introduce here the concept of VM plugin and interpreterproxy indirection/overhead. This is very important since later we just say this is a plugin and we don't redefine it}

@@

\paragraph{Final specification.} The final specification we productized last month takes two or three parameters. The third parameter, the order, is optional. If the third parameter is absent, the VM assumes the byte ordering (ASCII order) is the order to compare against. Therefore, the VM does not need to go through the indirection order array in the common case. The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.

In the paper we will focus on the comparison of two strings with the default ASCII ordering and measure the performance only of this case. The goal of this work was to improve the performance of common string operations, typically string equality which internally use the primitive, and the cases where the order is non ASCII are uncommon.

@@

\paragraph{Baseline: pure Smalltalk.}
To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk. There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings. We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the contructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level. The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:. Since the focus is only on ASCII ordered comparison, we show the code only of baselineCompareWith:.

\begin{code}
baselineCompareWith: aString 
	| c1 c2 length1 length2 |
	length1 := self size.
	length2 := aString size.
	1 to: (length1 min: length2) do: [ :i |
		(c1 := self basicAt: i) = (c2 := aString basicAt: i) ifFalse: [ ^ c1 - c2 ] ].
	^ length1 - length2
\end{code}

\paragraph{Existing: Smart Syntax Plugin.}

The Smart Syntax Plugin is one of the version existing until we started this work. The idea behind Smart Syntax was the following: the programmer can write code in Smalltalk using exclusively constructs that are recognized by the Smart Syntax compiler and that can be directly compiled to C (computing the size of a byte array, accessing its fields and doing integer arithmetic for example). With the addition of type annotations in the form of pragmas \cite{Duca16a} \footnote{Pragma are a form of method annotation.}, the Smart Syntax plugin was able to generate C code from the Smalltalk code. With this approach, the programmer has to write the optional primitive code a single time in Smalltalk. This code is used both as the fall-back code if the VM does not implement the primitive and at Slang to C compilation time to generate a C plugin from the code and the type annotations. This primitive follows the old specification.

\paragraph{Existing: Slang Plugin.}

The Slang plugin is a recent complete rewrite of the Smart Syntax plugin. The main difference is that the Smalltalk version is used only as fall-back code, while at Slang to C compilation time the VM uses code written in Slang to generate the C code. One of the side-effect is that the plugin's code is now part of the VM code base and not the client code bases, which is convenient for maintenance since the Cog VM has multiple client languages. However, the implementor now needs to maintain and evolve two version of the primitive, the fall-back code in Smalltalk and the code in Slang. This primitive follows the old specification.

@@

One or the other implementation could be chosen over another one, depending on what one's aiming at: the main trade-off will be between the engineering time and the performance. The table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation

@@

This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors. On the contrary, LLVM generates more instructions: it especially fails at perserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.

@@

%% Bibliography
\bibliography{sista,rmod}
